Informative Oceanographic Sampling using Excursion Probabilities for Multivariate Random Fields

### 2019-09-29 main_new.tex
Language updates Ginny

Adaptive sampling of an evolving frontal feature has been explored in
\cite{fronts11,Zhang2012,Pinto2018,costa19}. These approaches
typically use a reactive-adaptive scheme, whereby exploration does not
rely on a statistical model of the environment, but rather adapts
based on closing the sensing and actuation loop
(Fig. \ref{fig:sense-plan-act}). Myopic sampling—i.e. step-wise
selection of the path (on the graph) which reduces the information
criterion the most—has been used for adaptive surveys
\citep{singh2009efficient,Binney2013} that focus largely on reducing
predictive variance or mutual information (e.g. entropy). Variance and
entropy reduction are independent of the actual data realizations
under the assumptions of GP models; the use of data-driven adaptive
criteria was introduced to include more targeted sampling of regions
of scientific interest in \cite{Low2009} and \cite{fossuminformation}.
In this paper, with a focus on mapping the river plume, we reward
the designs that improve the classification of water masses as a
means to characterize the frontal zone.

@@

The ES holds random indicator variables while the associated EPs are: 
\begin{equation}\label{eq:prob}
  p(\bx) = P(\xi_T(\bx) \leq t_T, \xi_S(\bx) \leq t_S), \hspace{3mm} \bx \in \mathcal{M}.
\end{equation}
If the EPs in Eq. (\ref{eq:prob}) are
close to $1$ or $0$ at a given location, it is easy to classify water mass at this point as either river or ocean; it is, however,
challenging if the probability is close to $0.5$. 
The integrated Bernoulli variance (IBV) \cite{bect2019}, is defined as: 
\begin{equation}\label{measVAR}
    V = \int_{\bx \in \mathcal{M}} p(\bx) \left(1-p(\bx)\right) d\bx.
\end{equation}

@@

\subsection{Data model and updating of Gaussian processes}

@@

\subsection{Expected Integrated Bernoulli variance for bivariate models}

@@

A critical first element in the derivation is using the GP model and the
linear dependence on $\by$ in the conditional mean in
Eq. (\ref{gp_upd}). This results in a closed form Gaussian
distribution for the conditional mean in Eq. (\ref{dist_mxi}). The
high-dimensional inner integral in Eq. (\ref{sur}) then reduces to a
bivariate integral \citep{bhattacharjya2013value, chevalier2014fast},
with an expectation over $\bm=(m_{T},m_{S})^t$.  We standardize the two
variables in Eq. (\ref{sur}) as   $Z_T=\frac{\xi_T-m_{T}}{a_{T}}$,
$Z_S=\frac{\xi_S-m_{S}}{a_{S}}$; we can then rephrase the probability as
\begin{eqnarray}\label{ytom}
   P(\xi_T \leq t_T, \xi_S \leq t_S|\by) &=& P \left( Z_T \leq\frac{t_T-m_{T}}{a_{T}}, Z_S \leq\frac{t_S-m_{S}}{a_{S}}|\by \right) \nonumber \\
   &=& \Phi_2 \begin{pmatrix} 
\begin{bmatrix} \frac{t_T-m_{T}}{a_{T}}\\
\frac{t_S-m_{S}}{a_{S}}
\end{bmatrix};
 \begin{bmatrix} 0\\
0
\end{bmatrix},\begin{bmatrix}
1 & \eta_{s,t}  \\
\eta_{s,t}   & 1  
\end{bmatrix}\end{pmatrix} \nonumber \\
&=& P(\xi_T \leq t_T, \xi_S \leq t_S|\bm),
\end{eqnarray}
where $\eta_{T,S} =a_{T,S}/(a_{T} a_{S})$.
Since the expression only depends on $\by$ via its affine function $\bm$, the expectation in Eq. (\ref{sur}) is reduced to a bivariate integral over $\bm$. 
Introducing $\pi(\bm)$ for the density
function of $\bm$, the expected Bernoulli variance becomes:
\begin{eqnarray}\label{eq:var2}
E_{\by}(p(1-p)) &=& E_{\bm}(p(1-p)), \hspace{3mm} p=P(\xi_T \leq t_T, \xi_S \leq t_S|\bm) \nonumber \\
E_{\bm}(p(1-p)) & = & \int p[1-p] \pi(\bm) d\bm, \nonumber \\
 &=& \int P(\xi_T \leq t_T, \xi_S \leq t_S|\bm)  \pi(\bm) d\bm \nonumber  \\
&-& \int P(\xi_T \leq t_T, \xi_S \leq t_S|\bm) P(\xi_T \leq t_T, \xi_S \leq t_S|\bm) \pi(\bm) d\bm. 
\end{eqnarray}
We will next extend results of \cite{chevalier2014fast} to solve for
these two parts (see also \cite{stroh} for such an extensions).

@@

For the second part of the expression (\ref{eq:var2}), standardization
implies that: 
\begin{eqnarray}\label{part1:phi4}
&& \hspace{6mm} \int P(\xi_T \leq t_T, \xi_S \leq t_S|\bm) P(\xi_T \leq t_T, \xi_S \leq t_S|\bm) p(\bm) d\bm =  \\
&P& \hspace{-3mm} \left( Z_{1,T} \leq \frac{t_T-m_{T}}{a_{T}}, 
Z_{1,S} \leq \frac{t_S-m_{S}}{a_{S}},Z_{2,S} \leq \frac{t_T-m_{T}}{a_{T}}, 
Z_{2,S} \leq \frac{t_S-m_{S}}{a_{S}} \right). \nonumber
\end{eqnarray}
Here, $\bZ_1=(Z_{1,T},Z_{1,S})$, $\bZ_2=(Z_{2,T},Z_{2,S})$ are
independent bivariate zero-mean and unit-variance Gaussian vectors,
both with element-wise correlation $\eta_{T,S}$. Moreover, $\bZ_1$ and
$\bZ_2$ are chosen to be independent of $\bm$, hence:

@@

\subsection{Generalization to the multivariate case}
\label{gen_k}

@@

An important concept underlying this work is the potential of robotic sampling to adapt survey plans based on data and statistical inference
in accordance with the sense-plan-act loop
(Fig. \ref{fig:sense-plan-act}). We will therefore present a few adaptive strategies
for robotic sampling that build on the notion of multivariate ES. Simulations, analysis, and  results from full-scale trials are presented.

\subsection{Optimal sequential design}
\label{Optdes}

@@

At stage $s$, based on the currently available data
$\mathcal{Y}^{s-1}$, we fit an updated Gaussian model from
Eq. (\ref{gp_upd}) with mean $\bm_{s-1}=E(\bxi|\mathcal{Y}^{s-1})$
and covariance matrix $\bA_{s-1}=\mbox{Var}(\bxi|\mathcal{Y}^{s-1})$.
The next round of measurements $\by^{j,s}$ can be gathered with
design $d^{j,s}$, $j=1,\ldots,J$, where $J$ indicates all possible AUV
trajectories from the current stage. The {\it{naive}} strategy, then,
selects the design according to:
\begin{eqnarray}\label{critNaive}
    d^{*,s} &=& \mbox{argmin}_{j \in \{1,\ldots,J\}} |p(\bx_{d^{j,s}};\mathcal{Y}^{s-1})-0.5|, \\
    p(\bx;\mathcal{Y}^{s-1}) &=& P(\xi_T(\bx) \leq t_T, \xi_S(\bx) \leq t_S | \mathcal{Y}^{s-1}). \nonumber
\end{eqnarray}
This strategy does not account for the uncertainty in the temperature
or salinity variables, but is instead applicable only if one design (node) has EPs closer to
$0.5$ (see Table \ref{tab:sim_rhoab}, line two). Additionally, this strategy does not
account for spatial correlation. This strategy lacks memory of where
it has been and where the uncertainty has been reduced, and is therefore
susceptible to local minimas.

\subsection{Myopic path planning}
\label{myopic}

@@

\subsection{Look-ahead trajectory planning}
\label{LA}

@@

This look-ahead approach is much more computationally demanding than
the myopic strategy, so for practical implementation, we prune paths
in the evaluation of Eq. (\ref{critLA}). This means that we do not
compute all possible branches of the first two stages, as they are
indicated in Figure \ref{fig:PathSelOpt}. Instead, we use the myopic
strategy to rank the three best designs on the first stage alone, and
for each of these we undertake the look-ahead calculations.

@@

We now show results from tests performed to study the properties of
different static and sequential survey designs, in a realistic
simulated case. The context is mapping a freshwater plume defined by a
temperature and salinity gradient using an AUV.

@@

We use a bivariate GP model for temperature and salinity, where we
specify the prior mean:
\begin{equation}\label{m}
    \bmu(\bx)=E 
    \begin{bmatrix}
    \xi_T(\bx) \\
    \xi_S(\bx) 
    \end{bmatrix}=\begin{bmatrix} \mu_{T}(\bx)\\
\mu_{S}(\bx)
\end{bmatrix} 
= \begin{bmatrix} \beta_{T,0} + \beta_{T,1} x_{\mbox{East}} \\
\beta_{T,1} + \beta_{S,1} x_{\mbox{East}}
\end{bmatrix}.
\end{equation}
In this simulation setup, which is motivated by a real case, we expect
the eastern part of the domain to be cold freshwater. This situation
mimics that of a river mouth entering from the south, and the water
masses are pulled to the east. The actual trends are, in practice, set
from preliminary information or runs of oceanographic models \cite{fossuminformation} solving differential equations for hydrodynamic flow. In this
example the mean is specified by regression parameters $\beta_{T,1}=0.065$ and
$\beta_{S,1}=0.1$ for the slope with west coordinate and $\beta_{T,0}=5.8$
and $\beta_{S,1}=29.0$ for the intercept at a reference start map coordinate.

@@

The covariance is assumed to be stationary and separable for the two variables. 
The $2 \times 2$ pointwise covariance matrix is set to:
\begin{equation}\label{v0}
\bSigma(\bx)=\mbox{Var} 
\begin{bmatrix}
    \xi_T(\bx) \\
    \xi_S(\bx) 
    \end{bmatrix}=
\begin{bmatrix}
0.25^2 & 0.6 \cdot 0.25^2 \\
0.6 \cdot 0.25^2 & 0.25^2
\end{bmatrix}.
\end{equation}
The spatial correlation is of a Matern type:
\begin{equation}\label{v}
\mbox{Corr} 
\left(
\begin{bmatrix}
    \xi_T(\bx) \\
    \xi_S(\bx) 
    \end{bmatrix},
    \begin{bmatrix}
    \xi_T(\bx') \\
    \xi_S(\bx') 
    \end{bmatrix}
    \right)
    = \begin{bmatrix}
1 & 0.6  \\
0.6  & 1
\end{bmatrix}(1+\phi |\bx-\bx'|)\exp (-\phi |\bx-\bx'|),
\end{equation}
and $\phi=0.3$ indicates an effective correlation range of about $1200$ m. 
%Figure \ref{fig:stat_design} shows the contour lines for the EP for the reference. We notice the trend of increasing salinity and temperature to the west.  
We study sensitivity to the parameter specification in the results
below.  One realization of the random fields representing salinity and
temperature is shown in Figure \ref{fig:true_temp} and
\ref{fig:true_sal}. The true ES from this prior realization is shown
in Figure \ref{fig:ESet}.

@@

\subsection{Static and sequential sampling designs}\label{sec:sampling_designs}

@@

We compare three different static designs denoting
\texttt{static\_north}, \texttt{static\_east}, and
\texttt{static\_zigzag} with the three sequential approaches
\texttt{naive}, \texttt{myopic}, and \texttt{look-ahead}. The static
sampling paths are pre-determined and cannot be changed on the basis
of data, representing the typical pre-planned observation
strategies used in typical AUV operational survey designs.

For a fixed survey length, the expected variance is analytically available (equal coverage and reduction in variance each time). However, for the sequential approaches this is not the case, as the result and path are dependent on the observed data and the expected reduction in IBV. The properties is therefore evaluated using Monte Carlo sampling over several replicates of realizations from the prior, conducting simulated sequential surveying for each one. Figure \ref{fig:Eprob} shows the conditional EP, given data gathered along the north-south survey line for the realization indicated in these displays. In the Monte Carlo replicates, such results are repeatedly computed to approximate the expected variance reduction in the ES. We also compare predictive performance measured by root mean square error (RMSE) for the temperature and salinity estimates and the variance reduction in these two variables. It is important to note that the objective function used by the agents is focused on reducing the expected variance in the ES, but we nevertheless hope to achieve good predictive performance for criteria such as RMSE as well. Another non-statistical criteria that is important for practical purposes is the computational time of the strategy, as this will impact the performance of the approach in the field for an embedded system.

@@

Each strategy is conducted on an equilateral grid shown in Fig. \ref{fig:wp_graph}. The sequential sampling agents start at the center East-West coordinate at the southern end of the domain (node 53). The AUV will move along edges in the waypoint graph while the vehicle gathers measurements. The data is assimilated into the GP model, before an evaluation of the next node to sample is conducted at the end of the edge. The paths will differ for the various sequential designs due to the individual design criteria and the simulated variability in the environment among the replicates.

A total of 100 replicate simulation was conducted and the results are shown in Figure \ref{fig:sim_results},
where the different criteria are plotted as a function of survey distance. Figure \ref{fig:avg_ev} shows the resulting drop in ES variance for each of the six strategies. The ES variance is reduced most under the \texttt{myopic} and \texttt{look-ahead} strategies, performing almost equally; this is expected as the two criteria (Eq. \eqref{critSEQ} and \eqref{critLA}) are much more sensitive to differences in IBV. The north-south design also does well in terms of ES variance reduction since the path is parallel to the  boundary between the water masses.

@@

Figure \ref{fig:avg_rmse} and \ref{fig:avg_r2} show the resulting drop in RMSE and increase in explained variance. Both \texttt{myopic} and \texttt{look-ahead} perform well here, but some of the \texttt{static\_east} and \texttt{static\_zigzag} also achieve good results because they are pre-determined to cover large parts of the domain without re-visitation. Sequential strategies targeting ES variance will sometimes not reach similar coverage as interesting data may draw the path into turns and twists that may result in re-visitation.

\begin{figure}[!b]
  \centering
  \subfigure[Look-ahead]{\includegraphics[height = 0.46\textwidth]{Figures/sim/route_look-ahead.pdf}\label{fig:avg_look-ahead}}
  \hfill
  \subfigure[Myopic]{\includegraphics[height = 0.46\textwidth]{Figures/sim/route_myopic.pdf}\label{fig:avg_myopic}}
  \hfill
  \subfigure[Naive]{\includegraphics[height = 0.46\textwidth]{Figures/sim/route_naive.pdf}\label{fig:route_naive}}
  \hfill
  \subfigure[Static]{\includegraphics[height = 0.46\textwidth]{Figures/sim/route_static.pdf}\label{fig:route_static}}
  \hfill
  \caption{The overlaid route choices (strategies), superimposed for
    the 100 replicate surveys with 10 sampling choices/steps.}
\label{fig:route_choices}
\end{figure}

Fig.  \ref{fig:avg_time} shows the computational effort; the \texttt{naive} strategy is on par with the static designs, while the \texttt{myopic} strategy is slower and \texttt{look-ahead} more so, with the latter reaching levels that are nearly impractical for execution on a vehicle. Some pruning of the graph is performed to improve the computational speed, such as ruling out repeated visitations and back-and-forth routes. Some of the intermediate results are also stored for longer planning horizons. Still, more pruning of branches or inclusion of other heuristics could be included to make the strategy faster. Then again, the inclusion of heuristics is likely a contributing factor that the \texttt{look-ahead} strategy is not outperforming the \texttt{myopic} strategy. In Figure \ref{fig:route_choices} the realized sampling paths for each sequential schemes and static designs is shown.

The \texttt{naive} strategy gets stuck in the southern part of the domain in some cases, because it is too focused on the probabilities near $0.5$. The \texttt{myopic} strategy meanwhile, covers a wider domain than the look-ahead, there are several reasons for this. First, a greedy approach will tend to do more put more emphasis on locations close to the agent, which may lead away from the centre. Secondly, as the agent evaluates the impact of locations further away (look-ahead) where assimilated data has less effect, the prior (which is centered here) will act to restrict paths deviating from the central zone.

We further go on to study sensitivity of the result to different prior correlation values between temperature and salinity, prior standard deviations, spatial correlations, as well as measuring only temperature on-board the AUV (as indicated in the rows). We only compare results at the end of each survey. The expected ES variances are shown in Table \ref{tab:sim_res_ev}, results of RMSE are in Table \ref{tab:sim_res_rmse}, and explained variance in Table \ref{tab:sim_res_r2}. In all runs, both the \texttt{myopic} and \texttt{look-ahead} strategies perform the best in terms of expected variance in the ES. Surprisingly, the \texttt{myopic} is on-par and even slightly better than the \texttt{look-ahead} strategy. Part of the reason can be traced to the act of pruning some of the paths for the \texttt{look-ahead} strategy, but even with less pruning, the \texttt{look-ahead} does not outperform the \texttt{myopic} approach. The \texttt{static\_north} design is clearly better than the other two static designs, the reason being that the others do not focus on the areas where the ES is really uncertain (centre locations are the most uncertain). By measuring temperature alone the expected variances in the ES is only slightly higher. This occurs because the ES is almost the same and because of the correlation of $0.6$, which means that temperature observations provide information about salinity as well. For the other predictive measures, \texttt{myopic} and \texttt{look-ahead} do well, clearly beating the \texttt{naive} strategy. The \texttt{static\_zigzag} design is usually the best among the static designs for these predictive performance measures because its pattern gives more effective coverage of the domain when there is significant spatial correlation as in this case.

@@

\section{Case study - Mapping a river plume}
\label{sec:case_study}

@@

\subsection{Model specification}

The statistical model parameters were specified based on a short preliminary survey where the AUV did an initial transect to determine the environmental conditions and correlation structure. Based on the initial data the trend parameters were estimated by linear regression, where both temperature and salinity are assumed to increase linearly, going west from the river mouth. Next, the residuals from the regression analysis are analyzed to study the fit of the Gaussian model and to specify the covariance parameters.

>>>>

Adaptive sampling of an evolving frontal feature has been explored in
\cite{fronts11,Zhang2012,Pinto2018,costa19}. These approaches
typically use a reactive-adaptive scheme, whereby exploration does not
rely on a statistical model of the environment, but rather adapts
based on closing the sensing and actuation loop
(Fig. \ref{fig:envir1}). Myopic sampling—i.e. step-wise
selection of the path (on the graph) which reduces the information
criterion the most—has been used for adaptive surveys
\citep{singh2009efficient,Binney2013} that focus largely on reducing
predictive variance or mutual information (e.g. entropy). Variance and
entropy reduction are independent of the actual data realizations
under the assumptions of GP models; the use of data-driven adaptive
criteria was introduced to include more targeted sampling of regions
of scientific interest in \cite{Low2009} and \cite{fossuminformation}.
In this paper, with a focus on mapping the river plume, we reward
the designs that improve the classification of water masses as a
means to characterize the frontal zone.

@@

The ES holds random indicator variables while the associated EPs are: 
\begin{equation}\label{eq:prob}
  p(\bx) = P(\xi_T(\bx) \leq t_T, \xi_S(\bx) \leq t_S), \hspace{3mm} \bx \in \mathcal{M}.
\end{equation}
If the EPs in Eq. (\ref{eq:prob}) are
close to $1$ or $0$ at a given location, it is easy to classify water mass at this point as either river or ocean; it is, however,
challenging if the probability is close to $0.5$. 
The IBV \cite{bect2019}, is defined as: 
\begin{equation}\label{measVAR}
    V = \int_{\bx \in \mathcal{M}} p(\bx) \left(1-p(\bx)\right) d\bx.
\end{equation}

@@

\subsection{Data Model and Updating of Gaussian Processes}

@@

\subsection{Expected Integrated Bernoulli Variance for Bivariate Models}

@@

A critical first element in the derivation is using the GP model and the
linear dependence on $\by$ in the conditional mean in
Eq. (\ref{gp_upd}). This results in a closed form Gaussian
distribution for the conditional mean in Eq. (\ref{dist_mxi}). The
high-dimensional inner integral in Eq. (\ref{sur}) then reduces to a
bivariate integral \citep{bhattacharjya2013value, chevalier2014fast},
with an expectation over $\bm=(m_{T},m_{S})^t$.  We standardize the two
variables in Eq. (\ref{sur}) as   $Z_T=\frac{\xi_T-m_{T}}{a_{T}}$,
$Z_S=\frac{\xi_S-m_{S}}{a_{S}}$; we can then rephrase the probability as
\begin{eqnarray}\label{ytom}
   P(\xi_T \leq t_T, \xi_S \leq t_S|\by) &=& P \left( Z_T \leq\frac{t_T-m_{T}}{a_{T}}, Z_S \leq\frac{t_S-m_{S}}{a_{S}}|\by \right) \nonumber \\
   &=& \Phi_2 \begin{pmatrix} 
\begin{bmatrix} \frac{t_T-m_{T}}{a_{T}}\\
\frac{t_S-m_{S}}{a_{S}}
\end{bmatrix};
 \begin{bmatrix} 0\\
0
\end{bmatrix},\begin{bmatrix}
1 & \eta_{s,t}  \\
\eta_{s,t}   & 1  
\end{bmatrix}\end{pmatrix} \nonumber \\
&=& P(\xi_T \leq t_T, \xi_S \leq t_S|\bm),
\end{eqnarray}
where $\eta_{T,S} =a_{T,S}/(a_{T} a_{S})$.
Since the expression only depends on $\by$ via its affine function $\bm$, the expectation in Eq. (\ref{sur}) is reduced to a bivariate integral over $\bm$. 
Introducing $\pi(\bm)$ for the density
function of $\bm$, the expected Bernoulli variance becomes:
\begin{eqnarray}\label{eq:var2}
E_{\by}(p(1-p)) &=& E_{\bm}(p(1-p)), \hspace{3mm} p=P(\xi_T \leq t_T, \xi_S \leq t_S|\bm) \nonumber \\
E_{\bm}(p(1-p)) & = & \int p[1-p] \pi(\bm) d\bm, \nonumber \\
 &=& \int P(\xi_T \leq t_T, \xi_S \leq t_S|\bm)  \pi(\bm) d\bm \nonumber  \\
&-& \int P(\xi_T \leq t_T, \xi_S \leq t_S|\bm) P(\xi_T \leq t_T, \xi_S \leq t_S|\bm) \pi(\bm) d\bm. 
\end{eqnarray}
We will next extend results of \cite{chevalier2014fast} to solve for
these two parts (see also \cite{stroh}).

@@

For the second part of the expression Eq.(\ref{eq:var2}), standardization
implies that: 
\begin{eqnarray}\label{part1:phi4}
&& \hspace{6mm} \int P(\xi_T \leq t_T, \xi_S \leq t_S|\bm) P(\xi_T \leq t_T, \xi_S \leq t_S|\bm) p(\bm) d\bm =  \\
&P& \hspace{-3mm} \left( Z_{1,T} \leq \frac{t_T-m_{T}}{a_{T}}, 
Z_{1,S} \leq \frac{t_S-m_{S}}{a_{S}},Z_{2,S} \leq \frac{t_T-m_{T}}{a_{T}}, 
Z_{2,S} \leq \frac{t_S-m_{S}}{a_{S}} \right). \nonumber
\end{eqnarray}
Here, $\bZ_1=(Z_{1,T},Z_{1,S})$, $\bZ_2=(Z_{2,T},Z_{2,S})$ are
independent bivariate zero-mean and unit-variance Gaussian vectors,
both with element-wise correlation $\eta_{T,S}$. Moreover, $\bZ_1$ and
$\bZ_2$ are chosen to be independent of $\bm$, hence:

@@

\subsection{Generalization to the Multivariate Case}
\label{gen_k}

@@

An important concept underlying this work is the potential of robotic sampling to adapt survey plans based on data and statistical inference
in accordance with the sense-plan-act loop
(Fig. \ref{fig:envir1}). We will therefore present a few adaptive strategies
for robotic sampling that build on the notion of multivariate ES. Simulations, analysis, and  results from full-scale trials are presented.

\subsection{Optimal Sequential Design}
\label{Optdes}

@@

At stage $s$, based on the currently available data
$\mathcal{Y}^{s-1}$, we fit an updated Gaussian model from
Eq. (\ref{gp_upd}) with mean $\bm_{s-1}=E(\bxi|\mathcal{Y}^{s-1})$
and covariance matrix $\bA_{s-1}=\mbox{Var}(\bxi|\mathcal{Y}^{s-1})$.
The next round of measurements $\by^{j,s}$ can be gathered with
design $d^{j,s}$, $j=1,\ldots,J$, where $J$ indicates all possible AUV
trajectories from the current stage. The {\it{naive}} strategy, then,
selects the design according to:
\begin{eqnarray}\label{critNaive}
    d^{*,s} &=& \mbox{argmin}_{j \in \{1,\ldots,J\}} |p(\bx_{d^{j,s}};\mathcal{Y}^{s-1})-0.5|, \\
    p(\bx;\mathcal{Y}^{s-1}) &=& P(\xi_T(\bx) \leq t_T, \xi_S(\bx) \leq t_S | \mathcal{Y}^{s-1}). \nonumber
\end{eqnarray}
This strategy does not account for the uncertainty in the temperature
or salinity variables, but is instead applicable only if one design (node) has EPs closer to
$0.5$ (see Table \ref{tab:sim_rhoab}, line two). Additionally, this strategy does not
account for spatial correlation. This strategy lacks memory of where
it has been and where the uncertainty has been reduced, and is therefore
susceptible to local minima.

\subsection{Myopic Path Planning}
\label{myopic}

@@

\subsection{Look-ahead Trajectory Planning}
\label{LA}

@@

This look-ahead approach is much more computationally demanding than
the myopic strategy, so for practical implementation, we prune paths
in the evaluation of Eq. (\ref{critLA}). This means that we do not
compute all possible branches of the first two stages, as they are
indicated in Figure \ref{fig:PathSelOpt}. Instead, we use the myopic
strategy to rank the three best designs following the first stage alone, and
for each of these preferred designs, we undertake the look-ahead calculations.

@@

We now show results from tests performed to study the properties of
different static and sequential survey designs, in a realistic
simulated case: mapping a freshwater plume defined by a
temperature and salinity gradient using an AUV.

@@

We use a bivariate GP model for temperature and salinity, where we
specify the prior mean:
\begin{equation}\label{m}
    \bmu(\bx)=E 
    \begin{bmatrix}
    \xi_T(\bx) \\
    \xi_S(\bx) 
    \end{bmatrix}=\begin{bmatrix} \mu_{T}(\bx)\\
\mu_{S}(\bx)
\end{bmatrix} 
= \begin{bmatrix} \beta_{T,0} + \beta_{T,1} x_{\mbox{East}} \\
\beta_{T,1} + \beta_{S,1} x_{\mbox{East}}
\end{bmatrix}.
\end{equation}
In this simulation setup, which is motivated by a real case, we expect
the eastern part of the domain to be cold freshwater. This situation
mimics that of a river mouth opening from the south, with the water
masses pulled to the east. The actual trends are, in practice, set
from preliminary information or runs of oceanographic models \cite{fossuminformation} solving differential equations for hydrodynamic flow. In this
example, the mean is specified by regression parameters $\beta_{T,1}=0.065$ and
$\beta_{S,1}=0.1$ for the slope, with west coordinate and $\beta_{T,0}=5.8$
and $\beta_{S,1}=29.0$ for the intercept at a reference start map coordinate.

@@

The covariance is assumed to be stationary and separable for the two variables. 
The $2 \times 2$ pointwise covariance matrix is set to:
\begin{equation}\label{v0}
\bSigma(\bx)=\mbox{Var} 
\begin{bmatrix}
    \xi_T(\bx) \\
    \xi_S(\bx) 
    \end{bmatrix}=
\begin{bmatrix}
0.25^2 & 0.6 \cdot 0.25^2 \\
0.6 \cdot 0.25^2 & 0.25^2
\end{bmatrix}.
\end{equation}
The spatial correlation is of a Matern type:
\begin{equation}\label{v}
\small
\mbox{Corr} 
\left(
\begin{bmatrix}
    \xi_T(\bx) \\
    \xi_S(\bx) 
    \end{bmatrix},
    \begin{bmatrix}
    \xi_T(\bx') \\
    \xi_S(\bx') 
    \end{bmatrix}
    \right)
    = \begin{bmatrix}
1 & 0.6  \\
0.6  & 1
\end{bmatrix}(1+\phi |\bx-\bx'|)\exp (-\phi |\bx-\bx'|),
\end{equation}
and $\phi=0.3$ indicates an effective correlation range of about $1200$ m. 
%Figure \ref{fig:stat_design} shows the contour lines for the EP for the reference. We notice the trend of increasing salinity and temperature to the west.  
We study sensitivity to the parameter specification in the results
below.  One realization of the random fields representing salinity and
temperature is shown in Figures \ref{fig:true_temp} and
\ref{fig:true_sal}. The true ES from this prior realization is shown
in Figure \ref{fig:ESet}.

@@

\subsection{Static and Sequential Sampling Designs}\label{sec:sampling_designs}

@@

We compare three different static designs denoting
\textit{static\_north}, \textit{static\_east}, and
\textit{static\_zigzag} with the three sequential approaches
\textit{naive}, \textit{myopic}, and \textit{look-ahead}. The static
sampling paths are pre-determined and cannot be changed on the basis
of data, representing the pre-planned observation
strategies used in typical AUV operational survey designs.

For a fixed survey length, the expected variance is analytically available (equal coverage and reduction in variance each time). However, for the sequential approaches this is not the case, as the result and path are dependent on the observed data and the expected reduction in IBV. The properties are therefore evaluated using Monte Carlo sampling over several replicates of realizations from the prior, conducting simulated sequential surveying for each one. Figure \ref{fig:Eprob} shows the conditional EP, given data gathered along the north-south survey line for the realization indicated in these displays. In the Monte Carlo replicates, such results are repeatedly computed to approximate the expected variance reduction in the ES. We also compare predictive performance measured by root mean square error (RMSE) for the temperature and salinity estimates and the variance reduction in these two variables. It is important to note that the objective function used by the agents is focused on reducing the expected variance in the ES, but we nevertheless expect that we will achieve good predictive performance for criteria such as RMSE as well. Another non-statistical criteria that is important for practical purposes is the computational time of the strategy, as this will impact the performance of the approach in the field for an embedded system.

@@

Each strategy is conducted on an equilateral grid as shown in Fig. \ref{fig:wp_graph}. The sequential sampling agents start at the center East-West coordinate at the southern end of the domain (node 53). The AUV will move along edges in the waypoint graph while the vehicle gathers measurements. The data is assimilated into the GP model before an evaluation of the next node to sample is conducted at the end of the edge. The paths will differ for the various sequential designs due to the individual design criteria and the simulated variability in the environment among the replicates.

\subsection{Simulation Results}
A total of 100 replicate simulations were conducted; the results are shown in Figure \ref{fig:sim_results},
where the different criteria are plotted as a function of survey distance. Figure \ref{fig:avg_ev} shows the resulting drop in ES variance for each of the six strategies. The ES variance is reduced most under the \textit{myopic} and \textit{look-ahead} strategies, each performing almost equally; this is expected as the two criteria (Eq. \eqref{critSEQ} and \eqref{critLA}) are much more sensitive to differences in IBV. The north-south design also does well in terms of ES variance reduction since the path is parallel to the  boundary between the water masses.

@@

Figure \ref{fig:avg_rmse} and \ref{fig:avg_r2} show the resulting drop in RMSE and increase in explained variance, respectively. Both \textit{myopic} and \textit{look-ahead} strategies perform well here, but some of the \textit{static\_east} and \textit{static\_zigzag} also achieve good results because they are pre-determined to cover large parts of the domain without re-visitation. Sequential strategies targeting ES variance will sometimes not reach similar coverage, as interesting data may draw the path into twists and turns that may result in re-visitation.

\begin{figure}[!b]
  \centering
  \subfigure[Look-ahead]{\includegraphics[height = 0.46\textwidth]{Figures/sim/route_look-ahead.pdf}\label{fig:avg_look-ahead}}
  \hfill
  \subfigure[Myopic]{\includegraphics[height = 0.46\textwidth]{Figures/sim/route_myopic.pdf}\label{fig:avg_myopic}}
  \hfill
  \subfigure[Naive]{\includegraphics[height = 0.46\textwidth]{Figures/sim/route_naive.pdf}\label{fig:route_naive}}
  \hfill
  \subfigure[Static]{\includegraphics[height = 0.46\textwidth]{Figures/sim/route_static.pdf}\label{fig:route_static}}
  \hfill
  \caption{The overlaid route choices (strategies), superimposed on
    the 100 replicate surveys with 10 sampling choices/steps.}
\label{fig:route_choices}
\end{figure}

Fig.  \ref{fig:avg_time} shows the computational effort: the \textit{naive} strategy is on par with the static designs, while the \textit{myopic} strategy is slower. The \textit{look-ahead} is even slower, reaching levels that are nearly impractical for execution on a vehicle. Some pruning of the graph is performed to improve the computational speed, such as ruling out repeated visitations and back-and-forth routes. Some of the intermediate results are also stored for longer planning horizons. Still, more pruning of branches or inclusion of other heuristics could be included to make the strategy faster. Then again, the inclusion of heuristics is likely a contributing factor to the \textit{look-ahead} strategy failing to outperform the \textit{myopic} strategy. In Figure \ref{fig:route_choices}, the realized sampling paths for each of the sequential schemes and static designs is shown.

In this model, the \textit{naive} strategy often gets stuck in the southern part of the domain because it is too focused on the probabilities near $0.5$. The \textit{myopic} strategy, meanwhile, covers a wider domain than the naive or look-ahead. There are several reasons for this; first, a greedy approach will tend to do more put more emphasis on locations close to the agent, which may lead away from the centre. Second, as the agent evaluates the impact of locations further away (look-ahead) where assimilated data has less predictive power, the GP prior (which is centered here) will act to restrict paths deviating from the central zone.

\subsection{Performance Metrics}
The tables below explore the sensitivity of the result to different prior correlation values between temperature measured on-board the AUV, temperature and salinity, prior standard deviations, and spatial correlations (see table rows). We only compare results at the end of each survey. The expected ES variances are shown in Table \ref{tab:sim_res_ev}, results of RMSE are in Table \ref{tab:sim_res_rmse}, and explained variance is in Table \ref{tab:sim_res_r2}. In all runs, both the \textit{myopic} and \textit{look-ahead} strategies perform the best in terms of expected variance in the ES. Surprisingly, the \textit{myopic} performs slightly better than the \textit{look-ahead} strategy. Part of the reason for this performance discrepancy can be traced to the act of pruning some of the paths for the \textit{look-ahead} strategy; however, even with less pruning, the \textit{look-ahead} does not outperform the \textit{myopic} approach. The \textit{static\_north} design is clearly better than the other two static designs, the reason being that the others do not focus on the areas where the ES is really uncertain (centre locations are the most uncertain). By measuring temperature alone, the expected variances in the ES are only slightly higher than the combined temperature-salinity estimate. This occurs because the correlation of $0.6$ means that temperature observations provide information about salinity as well. For the other predictive measures, \textit{myopic} and \textit{look-ahead} strategies do well, clearly beating the \textit{naive} strategy. The \textit{static\_zigzag} design is usually the best among the static designs for these predictive performance measures because its pattern gives more effective coverage of the domain when there is significant spatial correlation, as indicated in this case.

@@

\section{Case Study - Mapping a River Plume}
\label{sec:case_study}

@@

\subsection{Model Specification}

The statistical model parameters were specified based on a short preliminary survey where the AUV did an initial transect to determine the environmental conditions and correlation structure. Based on the initial data, the trend parameters were estimated by linear regression, where both temperature and salinity are assumed to increase linearly, going west from the river mouth. Next, the residuals from the regression analysis were analyzed to study the fit of the Gaussian model and to specify the covariance parameters.
